Graph Topological Aspects of Granger Causal Network Learning

### 2019-07-12 paper_theory/paper_theory.tex
Breaking paper into separate pieces of work.  Also fixed significant error in one of the Lemmas.

\begin{document}
\maketitle
\abstract{We study Granger-causality in the context of wide-sense
  stationary time series, where our focus is on the topological
  aspects of the underlying causality graph.  We establish sufficient
  conditions (in particular, we develop the notion of a ``strongly
  causal'' graph topology) under which the true causality graph can be
  recovered via pairwise causality testing alone, and provide examples
  from the gene regulatory network literature suggesting that our
  concept of a strongly causal graph may be applicable to this field.
  We implement and detail finite-sample heuristics derived from our
  theory, and establish through simulation the efficiency gains (both
  statistical and computational) which can be obtained (in comparison
  to LASSO algorithms) when structural assumptions are met.  Finally,
  we provide an example application where we accurately discriminate
  between subjects in an EEG study based purely on the
  Granger-causality graphs inferred from data by our algorithms,
  demonstrating that meaningful features are captured by the
  Granger-causal graph topology.}

\keywords{causality graph, EEG, gene regulatory networks,
  Granger-causality, LASSO, network learning, time series, vector
  autoregression}

@@

\section{Introduction and Review}
\label{sec:introduction}
% Consider a collection of stochastic processes producing observations
% at discrete time intervals.  Are the underlying processes dependent?
% Can we quantify any of the underlying relationships?  Can the arrow of
% time help us to distinguish a directionality or flow of dependence
% among our observed series?  In this paper we contribute to the
% understanding of the notion of Granger-Causality
% \cite{granger1969investigating} \cite{Granger1980329} as a tool for
% answering these questions.

% Heuristically, we state that if an event '$A$' occuring strictly prior
% in time to an even '$B$' and $A$ provides us with exclusive (that is,
% not available anywhere else) information about $B$, then $A$ must have
% had a causal impact on $B$.  This is an entirely model-free notion of
% cause, and instead leverages the intuition that a cause must precede
% it's effect.

In this paper we study the notion of Granger-causality
\cite{granger1969investigating} \cite{Granger1980329} as a means of
uncovering an underlying causal structure in multivariate time series.
Though the underlying causality graph cannot be observed directly, we
will infer it's presence as a latent structure among our observed time
series data.  This notion is leveraged in a variety of applications
e.g. in Neuroscience as a means of recovering interactions amongst
brain regions \cite{bressler2011wiener}, \cite{anna_paper2008},
\cite{david2008identifying}; in the study of the dependence and
connectedness of financial institutions \cite{NBERw16223}; gene
expression networks \cite{Fujita2007},
\cite{methods_for_inferring_gene_regulatory_networks_from_time_series_expression_data},
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery},
\cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty};
and power system design \cite{Misyrlis2016450}, \cite{yuan2014root}.

@@

A second and wildly successful heuristic is the LASSO regularizer
\cite{tibshirani1996regression}, which can be understood as a natural
convex relaxation to penalizing the count of the non-zero edges.  The
LASSO enjoys fairly strong theoretical guarantees
\cite{wainwright2009sharp}, extending largely to the case of
stationary time series data with a sufficiently fast rate of
dependence decay \cite{basu2015} \cite{wong2016lasso}
\cite{autoregressive_process_modelling_via_the_lasso_procedure}, and
variations on the LASSO have been applied in a number of different
time series contexts as well as Granger-causality
\cite{DBLP:journals/corr/HallacPBL17} \cite{haufe2008sparse}
\cite{bolstad2011causal} \cite{he2013stationary}
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery}.
One of the key improvements to the original LASSO algorithm is the
adaptive (i.e. weighted) ``adaLASSO'' \cite{adaptive_lasso_zou2006},
for which oracle results (i.e. asymptotic support recovery) are
established under less restrictive conditions than for the vanilla
LASSO.  Our experimental comparisons in Section
\ref{sec:empirical_evaluation} are against the
adaLASSO. % Success of the LASSO has lead to
% the famous ``bet on sparsity'' principle ``Use a procedure that does
% well in sparse problems, since no procedure does well in dense
% problems.''  \cite{tibshirani2015statistical}.

\subsection{Contributions}
In the context of time series data, sparsity assumptions remain
important, but there is significant additional structure that may
arise as a result of considering the topology of the underlying
Granger-causality graph, which to our knowledge remains largely
unexplored.  The focus of this paper is to shed light on some of these
topological questions, in particular, we study a particularly simple
notion of causality graph topology which we term ``strongly causal''
and show that stationary times series whose underlying causality graph
has this structure satisfy natural intuitive notions of ``information
flow'' through the graph.  Moreover, we show that such graphs are
perfectly recoverable with only \textit{pairwise} Granger-causality
tests, which would otherwise suffer from serious confounding problems.
Our finite sample results are based on simulations, where we show
promising results for these graph topologies where our algorithm
performs substantially better in our simulation setup than do
competing LASSO algorithms, even for graphs that do not exactly
satisfy our strongly-causal topology assumptions.

@@

The principle contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger-causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as an interpretation in terms of the graph topology of the
sparsity pattern of matrices arising in the Wold decomposition,
generalizing in some sense the notion of ``feedback-free'' processes
studied by \cite{caines1975feedback} in close connection with
Granger-causality.  We establish sufficient conditions (sections
\ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
which a fully conditional Granger-causality graph can be recovered
from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).
Secondly, we propose in section \ref{sec:structure_learning} a graph
search heuristic which implements our theoretical results to finite
data samples, specifying and summarizing appropriate methods for
hypothesis testing (section \ref{sec:pairwise_hypothesis_testing}),
model order selection (section \ref{sec:model_order_selection}),
computationally efficient estimation (section
\ref{sec:efficient_model_estimation}), and error rate controls
(section \ref{sec:error_rate_control}).  Our heuristics are compared
against the adaLASSO algorithm in Section
\ref{sec:empirical_evaluation}.  We stress the scalability of our
algorithm which is capable of comfortably handling hundreds or
thousands of nodes on a single machine, as opposed to standard LASSO
algorithms which do not take advantage of the special structure
associated with stationary time series data.  In section
\ref{sec:application} we develop an example application where a
classifier is constructed to accurately discriminate between subjects
in an EEG study based only on the Granger-causality graph topology
inferred by our algorithms.  Concluding remarks on further open
problems and extentions are provided in Section \ref{sec:conclusion}.

\section{Theory}
\label{sec:theory}
\subsection{Formal Setting}
Consider the space $L_2(\Omega)$, the usual Hilbert space of finite
variance random variables over a probability space
$(\Omega, \mathcal{F}, \mathbb{P})$ having inner product
$\inner{x}{y} = \E[xy]$.  We will work with a discrete time and
wide-sense stationary (WSS) $n$-dimensional vector valued process
$x(t)$ (with $t \in \Z$) where the $n$ elements take values in $L_2$.  We
suppose that $x(t)$ has zero mean, $\E x(t) = 0$, and has absolutely
summable matrix valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$ with an absolutely
continuous spectral density $S(\omega)$:

\begin{equation}
  \label{eqn:fourier_pair}
  \begin{aligned}
    R(\tau) &= \frac{1}{2\pi}\int_{-\pi}^\pi S(\omega) e^{j\tau\omega}\d \omega,\\
    S(\omega) &= \sum_{\tau=-\infty}^\infty R(\tau)e^{-j\tau\omega}.
  \end{aligned}
\end{equation}

% and that the spectra is bounded uniformly away from 0:

% \begin{equation*}
%   \int_{-\pi}^\pi \ln\det S(\omega) \d\omega > -\infty.
% \end{equation*}

@@

where $c(t)$ is a purely deterministic sequence\footnote{the purely
  deterministic sequence $c(t)$ is one which lies in the remote past
  $\bigcap_{\tau=1}^\infty \H_{t - \tau}^x$ of the process.  For such
  processes a single sample $c(t_0)$ is enough to determine $c(t)$ for
  every $t$.  For example,
  $c(t) = \text{sin}(2\pi t + \Theta);\; \Theta \sim \mathcal{U}[-\pi,
  \pi]$}, $v(t)$ is an uncorrelated sequence and $A(0) = I$.  We will
assume that $c(t) = 0$, which in practice is to say that the process
has been detrended.  We additionally require that this representation
can be inverted to yield the $\VAR(\infty)$ form

@@

Finally, since $||z^{-1}|| = 1$ we have the inversion formula

@@

The aforementioned assumptions are quite weak.  The strongest
assumption we require is finally that $\Sigma_v$ is a diagonal matrix,
which is referred to as a lack of instantaneous feedback in $x(t)$.
We formally state our setup to conclude this section:

@@

This notion captures the idea that the process $x_j$ provides
information about $x_i$ that is not available from elsewhere.  The
caveat ``with respect to $x$'' is important in that GC relations can
change when components are added to or removed from our collection $x$
of observations, e.g. new GC relations can arise if we remove the
observations of a common cause, and existing GC relations can
disappear if we observe a new mediating series.

The notion is closely related to the information theoretic measure of
transfer entropy, indeed, if the distribution of $v(t)$ is known to be
Gaussian then they are equivalent \cite{barnett2009granger}.

We require two technical facts we refer to in the sequel, the notion
of conditional orthogonality is used throughout.

@@

@@

    \begin{equation*}
      \inner{a - \linE{a}{\mathcal{X}}}{b - \linE{b}{\mathcal{X}}} = 0\ \forall a \in \mathcal{A}, b \in \mathcal{B}.
    \end{equation*}
\end{lemma}
% \begin{proof}
%   Proceding from the definition of conditional orthogonality we have

%   \begin{align*}
%     &\inner{\alpha - \linE{\alpha}{\mathcal{X}}}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
%     &{\iff} \inner{\alpha + x - \linE{\alpha + x}{\mathcal{X}}}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
%     &\overset{(a)}{\iff} \inner{\alpha + x}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
%     &\overset{(b)}{\iff} \linE{\beta - \linE{\beta}{\mathcal{X}}}{\mathcal{A} \vee \mathcal{X}} = 0\ \forall \beta \in \mathcal{B}\\
%     &\overset{(c)}{\iff} \linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}},
%   \end{align*}

%   where $(a)$ follows since $(\beta - \linE{\beta}{\mathcal{X}}) \perp \mathcal{X}$ and $\linE{\alpha + x}{\mathcal{X}} \in \mathcal{X}$, $(b)$ since $z \perp \mathcal{X} \vee \mathcal{A} \iff \linE{z}{\mathcal{X} \vee \mathcal{A}} = 0$, and $(d)$ since $\linE{z}{\H_2} = \linE{\linE{z}{\H_2}}{\H_1}$ for $\H_2 \subseteq \H_1$.

%   % This is property (vi) in Lindquist
%   % \begin{align*}
%   %   \langle \alpha - \linE{\alpha}{\mathcal{X}}, \beta - \linE{\beta}{\mathcal{X}} \rangle &= 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
%   %   \overset{(a)}{\iff} \langle \alpha, \beta - \linE{\beta}{\mathcal{X}} \rangle &= 0\ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
%   %   \overset{(b)}{\iff} \linE{\beta - \linE{\beta}{\mathcal{X}}}{\mathcal{A}} &= 0\ \forall \beta \in \mathcal{B}
%   % \end{align*}

%   % where $(a)$ follows because $\linE{\alpha}{\mathcal{X}} \in \mathcal{X}$ and $\beta - \linE{\beta}{\mathcal{X}} \in \mathcal{X}$, and $(b)$ follows since $\beta \perp \mathcal{A} \iff \linE{\beta}{\mathcal{A}} = 0$.
% \end{proof}
\begin{lemma}
  \label{lem:subspace_sum_projection}
  Let $x \in \H$ where $\H$ is a separable Hilbert space having inner product
  $\inner{\cdot}{\cdot}$ and let $\{\H_n\}_{n = 0}^N$ be a
  collection of subspaces of $\H$ such that $x \perp \H_0$.  Then

  \begin{equation*}
    \linE{x}{\bigvee_{n = 0}^N\H_n} = \linE{x}{\bigvee_{n = 1}^N\H_n},
  \end{equation*}

  where
  $\H_1 \vee \H_2 = \cl \{\alpha + \beta\ |\ \alpha \in \H_1, \beta
  \in \H_2 \}$ is the closed sum of subspaces.
\end{lemma}
\begin{proof}
  Apply the Gram-Schmidt process and directly calculate the projections.
\end{proof}

\begin{theorem}[Granger Causality Equivalences]
  \label{thm:granger_causality_equivalences}
  Let $x(t)$ be a WSS process with absolutely summable covariance
  sequence, and spectral density uniformly bounded above and below.
  Denote by $\xi_{ij} \defeq \linEerr{x_i(t)}{\H^{-j}}$ and
  $\xi_i \defeq \linEerr{x_i(t)}{\H_t}$ for distinct $i, j$.  Then,
  the following are equivalent:

  \begin{enumerate}
    \item{$x_j \ngc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$ i.e. $\B_{ij}(z) = 0$}
    \item{$F_{ij} \defeq \big(\frac{\xi_i}{\xi_{ij}} - 1\big) = 0$}
    \item{$\H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j} \iff H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j}$}
    \item{$\linE{x_i(t)}{\H_{t - 1}^{-j}} = \linE{x_i(t)}{\H_{t - 1}}$}
    % \item{\hl{Wold $A(\tau)$ condition}}
    % \item{Tie in the 0s of $S(\omega)^{-1}$, though this isn't immediately directed.}
  \end{enumerate}
\end{theorem}

@@

  $(1) \iff (3)$ is a tautology, and $(3) \iff (5)$ is immediate from the
  definitions i.e. the projections are equal if and only if the errors
  are equal.  We have $(4) \iff (5)$ as a result of Lemma
  \ref{lem:conditional_orthogonality_equivalence}

% , the additional
%   equivalence in $(4)$ is apparent since $\H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$.

  $(1) \Rightarrow (2)$: The projection for $x_i(t)$ onto $\H_{t - 1}$ is
  (by definition) given by a form similar to equation
  \ref{eqn:ar_representation}:

\[
  \E|x_i(t) - \linE{x_i(t)}{\H_{t - 1}}|^2 = \E|v_i(t) + \sum_{\tau = 1}^\infty \sum_{k = 1}^n (B_{i, k}(\tau) - \hat{B}_{i, k}(\tau))x_k(t - \tau)|^2.
\]

Since $v(t)$ in equation \ref{eqn:ar_representation} is temporally
uncorrelated, it follows that the optimal projection is given by the
model coefficients themselves.  This holds similarly for the
projection onto $\H_{t - 1}^{-j}$.  Then by $(1)$ we have

\[
  \E|x_i(t) - \sum_{\tau = 1}^\infty \sum_{k = 1}^n B_{i, k}(\tau)x_k(t - \tau)|^2 = \E|x_i(t) - \sum_{\tau = 1}^\infty\sum_{k \ne j}B_{i, k}(\tau)x_k(t - \tau)|^2
\]

By the uniqueness of the projection we must have
$\forall \tau\ B_{i, j}(\tau) = 0$.

$(2) \Rightarrow (4)$: In computing $(y - \linE{y}{\H_{t - 1}^{-j}})$ for
$y \in H_t^i$ it is sufficient to consider $y = x_i(t)$ by linearity, then since
$H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$ we have
$(x_i(t) - \linE{x_i(t)}{\H_{t = 1}^{-j}}) = v_i(t)$ since $\B_{ij}(z) = 0$.  $(4)$ then follows
since $v_i(t) \perp \H_{t - 1}$ and
$\forall z \in \H_{t - 1}^j\ (z - \linE{z}{\H_{t - 1}^{-j}}) \in \H_{t
  - 1}$
\end{proof}

% The following propositions justify various modifications of $x(t)$
% applied in practice to massage $x(t)$ into a form amenable to more
% standard tools.

% \begin{theorem}[General Invariance]
%   \hl{Under what conditions is this actually true?}

%   Let $\zeta(t)$ be a stationary discrete time stochastic process.  Let
%   $F, G$ be (possibly nonlinear and time varying) invertible filtering
%   operations and $f(t)$, $g(t)$ be perfectly predictable and such that
%   $x_j(t) \defeq F(\zeta_j - f)(t), x_i(t) \defeq G(\zeta_i - g)(t)$ are W.S.S.  Then,

% \begin{equation}
%     x_j \gc x_i \iff \zeta_j \te \zeta_i
%   \end{equation}
% \end{theorem}

% \begin{proposition}[Invariance Under Invertible and Deterministic Modifications]
%   Let $x(t)$ be a WSS process with absolutely summable covariance
%   sequence, and spectral density uniformly bounded above and below.  Let
%   $F(z), G$ represent univariate and invertible
%   linear-time-invariant filters.  And, let $f(t), g(t)$ be perfectly
%   predictable processes.  Then,

%   \begin{equation}
%     x_j \gc x_i \iff F(x_j - f)(t) \gc G(x_i - g)(t)
%   \end{equation}

% \begin{proof}
%   \hl{TODO}
% \end{proof}
% \end{proposition}

\subsection{Granger Causality Graphs}
We first need to establish some graph theoretic notation and
terminology, collected formally in definitions for the reader's
convenient reference.

@@

  Recall that we do not allow a node to be it's own parent, however
  unless $\gcg$ is a DAG, a node can be it's own ancestor.  We will
  ocassionally need to explicitly exclude $i$ from $\anc{i}$, in which
  case we will write $\anc{i}\setminus \{i\}$.
\end{definition}

@@

% What if we take this as the definition of PWGC?  then, in a simply causal graph
% the "normal" definition becomes a theorem.  Or, if lem:pwgc_anc is true then
% maybe that is good enough?

@@

% % When is this guaranteed w.r.t. the full system?
% Furthermore, if the filter $\B_{ii}$ is invertible, we can write

% \begin{equation*}
%   \begin{aligned}
%     (1 - \B_{ii}(z))x_i(t) &= v_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t)\\
%     \Rightarrow x_i(t) &= (1 - \B_{ii}(z))^{-1}\big[v_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t)\big]\\
%     &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)x_k(t)
%   \end{aligned}
% \end{equation*}

% In general, we will write $\wtB$ for arbitrary filters maintaining the
% convention that $\wtB_{ii}$ is a filter that modifies the future of
% $x_i$ from it's past and $\wtB_{ij}$ a filter modifying $x_i$ from the
% past of $x_j$.  We will use the convention that the filter
% $\widetilde{\B}$ will represent simply ``a filter'' and is not
% necessarily the same as the filters serving as edge weights in $\gcg$
% (indeed, $\wtB$ will often be $0$), and moreover, we will allow for
% $\widetilde{\B}$ to potentially change from place to place and even
% from line to line in the same sequence of calculations as it's exact
% properties or specification are not important.

% We continue to develop the relationship between $x_i$ and it's family
% tree, where the following proposition is analagous to the definitions
% of feedback free processes of \cite{caines1975feedback}.

@@

% TODO: ???
% , however the implication $j \in \anc{i} \implies \A_{ij}(z) \ne 0$ need not hold.

% \begin{proof}
%   The formula (\ref{eqn:parent_expansion}) was established in the
%   preceding paragraph and serves as a starting point for the induction
%   hypothesis:

%   \begin{equation}
%     \label{eqn:induction}
%     x_i(t) = \wtB_{ii}(z)v_i(t) + \sum_{k \in \ancn{\ell - 1}{i}\setminus \{i\}}\wtB_{ik}(z)v_k(t) + \sum_{k \in \gpn{\ell}{i} \setminus \{i\}}\wtB_{ik}(z)x_k(t).
%   \end{equation}

%   The base case follows by expanding each $x_k$ in equation \ref{eqn:induction} via equation \ref{eqn:parent_expansion}, but we skip the calculations since they are nearly identical to the following induction step:

%   \begin{align*}
%     x_i(t) &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \ancn{\ell - 1}{i}\setminus \{i\}}\wtB_{ik}(z)v_k(t) + \sum_{k \in \gpn{\ell}{i} \setminus \{i\}}\wtB_{ik}(z)x_k(t)\\
%            &\overset{(a)}{=} \wtB_{ii}(z)v_i(t) + \sum_{k \in \ancn{\ell - 1}{i}\setminus \{i\}}\wtB_{ik}(z)v_k(t) + \sum_{k \in \gpn{\ell}{i} \setminus \{i\}}\wtB_{ik}(z)\big[\wtB_{kk}v_k(t)\\
%     &\ \ \ \  + \wtB_{ki}(z)x_i(t) + \sum_{h \in \pa{k}\setminus \{i\}}\wtB_{kh}(z)x_h(t)\big]\\
%     &\overset{(b)}{=} \wtB_{ii}(z)v_i(t) + \sum_{k \in \ancn{\ell}{i} \setminus \{i\}}\wtB_{ik}(z)v_k(t) + \sum_{k \in \gpn{\ell + 1}{i} \setminus \{i\}}\wtB_{ik}(z)x_k(t).
%   \end{align*}

%   Where equality $(a)$ follows by expanding each $x_k$ with equation \ref{eqn:parent_expansion} and $(b)$ requires the inversion of $\big(1 - \sum_{k \in \gpn{\ell}{i}\setminus\{i\}}\wtB_{ik}(z)\wtB_{ki}(z)\big)$ as well as using the fact that $\ancn{\ell}{i} = \ancn{\ell - 1}{i}\cup\gpn{\ell}{i}$.  We here remind the reader of our earlier established convention that $\wtB$ simply represents the existence of a filter (possibly 0) and can change from place to place.

% The induction terminates after $n$ steps since in a graph with $n$ nodes $\gpn{n + 1}{i} = \emptyset$

% % Where step $(a)$ requires the inversion of $1 - \sum_{k \in \pa{i}}\wtB_{ik}(z)\wtB_{ki}(z)$.

% %   \begin{align*}
% %     x_i(t) &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)x_k(t)\\
% %            &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)\big[\wtB_{kk}v_{k}(t) + \wtB_{ki}(z)x_i(t) + \sum_{h \in \pa{k}\setminus \{i\}}\wtB_{kh}(z)x_h(t) \big]\\
% %            &\overset{(a)}{=} \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)v_{k}(t) + \sum_{k \in \gpn{2}{i} \setminus \{i\}}\wtB_{ik}(z)x_k(t)
% %   \end{align*}

% % Where step $(a)$ requires the inversion of $1 - \sum_{k \in \pa{i}}\wtB_{ik}(z)\wtB_{ki}(z)$.

% \end{proof}
% \textbf{this is an alternative proof, I think superior to the one above}.

@@

  \begin{align*}
    x_i(t) &= [(I - \B(z))^{-1}v(t)]_i\\
    &= \sum_{j = 1}^n \A_{ij}(z) v_j(t)\\
    &= \A_{ii}(z) v_i(t) + \sum_{\substack{j \in \anc{i} \\ j \ne i}} \A_{ij}(z) v_j(t)
  \end{align*}
\end{proof}

@@

This notion is of interest for a variety of reasons.  From a purely
conceptual standpoint, we will see how the notion can in some sense
capture the idea of ``flow of information'' in the underlying graph,
in the sense that if $j \in \anc{i}$ we expect that $j \pwgc i$.  It
may also be useful for reasoning about the conditions under which
\textit{unobserved} components of $x(t)$ may or may not interfere with
inference in the actually observed components.  Finally, motivated
from a practical standpoint to analyze causation in large systems, we
will seek to construct practical estimation procedures based purely on
pairwise causality tests since the computation of such pairwise
relations is somewhat easier.

The following is a fundamental building block:

\begin{lemma}
  \label{lem:ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger-causality graph
  $\gcg$.  Then, if

>>>>

\begin{document}
\maketitle
\abstract{We study Granger-causality in the context of wide-sense
  stationary time series, where our focus is on the topological
  aspects of the underlying causality graph.  We establish sufficient
  conditions (in particular, we develop the notion of a ``strongly
  causal'' graph topology) under which the true causality graph can be
  recovered via pairwise causality testing alone, and provide examples
  from the gene regulatory network literature suggesting that our
  concept of a strongly causal graph may be applicable to this field.}

\keywords{causality graph, gene regulatory networks,
  Granger-causality, network learning, time series, vector
  autoregression}

@@

\section{Introduction and Review}
\label{sec:introduction}
In this paper we study the notion of Granger-causality
\cite{granger1969investigating} \cite{Granger1980329} as a means of
uncovering an underlying causal structure in multivariate time series.
Though the underlying causality graph cannot be observed directly, we
will infer it's presence as a latent structure among our observed time
series data.  This notion is leveraged in a variety of applications
e.g. in Neuroscience as a means of recovering interactions amongst
brain regions \cite{bressler2011wiener}, \cite{anna_paper2008},
\cite{david2008identifying}; in the study of the dependence and
connectedness of financial institutions \cite{NBERw16223}; gene
expression networks \cite{Fujita2007},
\cite{methods_for_inferring_gene_regulatory_networks_from_time_series_expression_data},
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery},
\cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty};
and power system design \cite{Misyrlis2016450}, \cite{yuan2014root}.

@@

A second and wildly successful heuristic is the LASSO regularizer
\cite{tibshirani1996regression}, which can be understood as a natural
convex relaxation to penalizing the count of the non-zero edges.  The
LASSO enjoys fairly strong theoretical guarantees
\cite{wainwright2009sharp}, extending largely to the case of
stationary time series data with a sufficiently fast rate of
dependence decay \cite{basu2015} \cite{wong2016lasso}
\cite{autoregressive_process_modelling_via_the_lasso_procedure}, and
variations on the LASSO have been applied in a number of different
time series contexts as well as Granger-causality
\cite{DBLP:journals/corr/HallacPBL17} \cite{haufe2008sparse}
\cite{bolstad2011causal} \cite{he2013stationary}
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery}.
One of the key improvements to the original LASSO algorithm is the
adaptive (i.e. weighted) ``adaLASSO'' \cite{adaptive_lasso_zou2006},
for which oracle results (i.e. asymptotic support recovery) are
established under less restrictive conditions than for the vanilla
LASSO.  We study a novel perspective on the problem by establishing
conditions under which simple \textit{pairwise} Granger-causality
tests are sufficient to recover the entire graph.

%   Prior assumptions
% on the \textit{topology} of the underlying Granger-causality graph can
% result in tests more powerful than more generally applicable
% algorithms, and we provide examples from the literature where
% simplifying topological assumptions may be met.

\subsection{Contributions}
In the context of time series data, sparsity assumptions remain
important, but there is significant additional structure that may
arise as a result of considering the topology of the underlying
Granger-causality graph, which to our knowledge remains largely
unexplored.  The focus of this paper is to shed light on some of these
topological questions, in particular, we study a particularly simple
notion of causality graph topology which we term ``strongly causal''
and show that stationary times series whose underlying causality graph
has this structure satisfy natural intuitive notions of ``information
flow'' through the graph.  Moreover, we show that such graphs are
perfectly recoverable with only \textit{pairwise} Granger-causality
tests, which would otherwise suffer from serious confounding problems.
Aside from being an interesting theoretical perspective, prior
assumptions about the underlying graph (similarly to sparsity
assumptions) can greatly improve upon the statistical power of
causality graph recovery algorithms when the assumptions are met.

@@

The principle contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger-causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as an interpretation in terms of the graph topology of the
sparsity pattern of matrices arising in the Wold decomposition,
generalizing in some sense the notion of ``feedback-free'' processes
studied by \cite{caines1975feedback} in close connection with
Granger-causality.  We establish sufficient conditions (sections
\ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
which a fully conditional Granger-causality graph can be recovered
from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).  We
report \hl{elsewhere}\todo{where?  Or can I fit in a brief summary?}
the results of extensive finite sample simulations as well as an
application to EEG data, where in addition we summarize appropriate
methods for hypothesis testing, model order selection, computationally
efficient estimations, and error rate controls in practice.  Our
simulation results establish significant potential for improvement
over existing methods.  Concluding remarks on further open problems
and extentions are provided in Section \ref{sec:conclusion}.

\section{Graph Topological Aspects of Granger-causality}
\label{sec:theory}
\subsection{Formal Setting}
Consider the space $L_2(\Omega)$, the usual Hilbert space of finite
variance random variables over a probability space
$(\Omega, \mathcal{F}, \mathbb{P})$ having inner product
$\inner{x}{y} = \E[xy]$.  We will work with a discrete time and
wide-sense stationary (WSS) $n$-dimensional vector valued process
$x(t)$ (with $t \in \Z$) where the $n$ elements take values in $L_2$.
We suppose that $x(t)$ has zero mean, $\E x(t) = 0$, and has
absolutely summable matrix valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$, and an absolutely continuous
spectral density.

@@

where $c(t)$ is a purely deterministic sequence$v(t)$ is an
uncorrelated sequence and $A(0) = I$.  We will assume that $c(t) = 0$.
We additionally require that this representation can be inverted to
yield the $\VAR(\infty)$ form

@@

Finally, we have the inversion formula

@@

The aforementioned assumptions are quite weak.  The strongest
assumption we require is finally that $\Sigma_v$ is a diagonal matrix,
which is referred to as a lack of instantaneous feedback in $x(t)$.
We formally state our setup as a definition, which is the setup for
the remainder of the paper:

@@

This notion captures the idea that the process $x_j$ provides
information about $x_i$ that is not available from elsewhere.  The
caveat ``with respect to $x$'' is important in that GC relations can
change when components are added to or removed from our collection $x$
of observations, e.g. new GC relations can arise if we remove the
observations of a common cause, and existing GC relations can
disappear if we observe a new mediating series. The notion is closely
related to the information theoretic measure of transfer entropy,
indeed, if the distribution of $v(t)$ is known to be Gaussian then
they are equivalent \cite{barnett2009granger}.

The notion of conditional orthogonality is used throughout, as well as
some equivalent representations for Granger-causality.

@@

  \todo{Do I even use (a) anymore?  I think I can just put the definition of conditional orthogonality into Theorem 1.}

@@

    \begin{equation*}
      \inner{a - \linE{a}{\mathcal{X}}}{b - \linE{b}{\mathcal{X}}} = 0\ \forall a \in \mathcal{A}, b \in \mathcal{B}.
    \end{equation*}
\end{lemma}

\begin{theorem}[Granger Causality Equivalences]
  \label{thm:granger_causality_equivalences}
  The following are equivalent:

  \begin{enumerate}
    \item{$x_j \ngc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$ i.e. $\B_{ij}(z) = 0$}
    \item{$H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j}$}
    \item{$\linE{x_i(t)}{\H_{t - 1}^{-j}} = \linE{x_i(t)}{\H_{t - 1}}$}
  \end{enumerate}
\end{theorem}

@@

  $(1) \Rightarrow (2)$ follows as a result of the uniqueness of orthogonal
  projection (i.e. the best estimate is necessarily the coefficients
  of the model).  $(2) \Rightarrow (3)$ follows since in computing
  $(y - \linE{y}{\H_{t - 1}^{-j}})$ for $y \in H_t^i$ it is sufficient
  to consider $y = x_i(t)$ by linearity, then since
  $H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$ we have
  $(x_i(t) - \linE{x_i(t)}{\H_{t = 1}^{-j}}) = v_i(t)$ since
  $\B_{ij}(z) = 0$ and $v_i(t) \perp \H_{t - 1}$.  $(3) \iff (4)$ is a result
  of Lemma \ref{lem:conditional_orthogonality_equivalence}.  And,
  $(4) \implies (1)$ follows directly from the Definition.
\end{proof}

\subsection{Granger Causality Graphs}
We establish some graph theoretic notation and terminology, collected
formally in definitions for the reader's convenient reference.

@@

  Recall that we do not allow a node to be it's own parent, however
  unless $\gcg$ is a DAG a node \textit{can} be it's own ancestor.  We will
  ocassionally need to explicitly exclude $i$ from $\anc{i}$, in which
  case we will write $\anc{i}\setminus \{i\}$.
\end{definition}

@@

@@

@@

@@

  \begin{align*}
    x_i(t) &= [(I - \B(z))^{-1}v(t)]_i = \sum_{j = 1}^n \A_{ij}(z) v_j(t)\\
    &= \A_{ii}(z) v_i(t) + \sum_{\substack{j \in \anc{i} \\ j \ne i}} \A_{ij}(z) v_j(t)
  \end{align*}
\end{proof}

@@

This notion is of interest for a variety of reasons.  From a purely
conceptual standpoint, we will see how the notion can in some sense
capture the idea of ``flow of information'' in the underlying graph,
in the sense that if $j \in \anc{i}$ we expect that $j \pwgc i$.  It may
also be useful for reasoning about the conditions under which
\textit{unobserved} components of $x(t)$ may or may not interfere with
inference in the actually observed components.  Finally, motivated
from a practical standpoint to analyze causation in large systems,
practical estimation procedures based purely on pairwise causality
tests are of interest since the computation of such pairwise relations
is substantially easier..

\begin{lemma}
  \label{lem:separated_ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger-causality graph
  $\gcg$.  If

  \begin{enumerate}[label=(\alph*)]
    \item{$j \not\in \anc{i}$ and $i \not\in \anc{j}$}
    \item{$\anc{i}\cap\anc{j} = \emptyset$}
  \end{enumerate}
